hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-01-07 inst/manuscript/paper.Rmd
discussion

In the application of using effective value models **hatchR** and the user make some key assumptions. Particularly, numerous studies have indicated that stressful environmental conditions can cue fish to prematurely hatch or emerge from their environment. These include water quality like dissolved oxygen, pH, or salnity, pathogens, and even mechanical agitation (reviewed in @quinn2018 and @cowan2024). Moreover, while the **hatchR** provides point estimates for developmental phenology, spawning and developmental within populations of fishes both occur as distributions as opposed single events @mason1976 and we encourage users to either predict phenology using early, peak, and late frameworks (*e.g.*, 5th, 50th, and 95th quantiles) or with real or modeled distributions.

@@

talk about how these models represent point estimates and that emergence and hatch will take the form of a distribution

talk about how temperatures may not always reflect redd temperture

talk about how other factors like oxygen or predator cues may change phenology

talk about how these models could be extended for other development phases or to organisms that fit similar developmental patterns

talk about how advanced users could fit their own models and provide their custom expression to predict_phenology for a really customized option.

>>>>

In the application of using effective value models **hatchR** and the user make some key assumptions. Particularly, numerous studies have indicated that stressful environmental conditions can cue fish to prematurely hatch or emerge from their environment. These include water quality like dissolved oxygen, pH, or salnity, pathogens, and even mechanical agitation (reviewed in @quinn2018 and @cowan2024). Moreover, while the **hatchR** provides point estimates for developmental phenology, spawning and developmental within populations of fishes both occur as distributions as opposed single events @mason1976 and we encourage users to either predict phenology using early, peak, and late frameworks (*e.g.*, 5th, 50th, and 95th quantiles) or with real or modeled distributions. Another factor that could bias estimates from these models is that temperatures used from sensors don't reflect the geomoprhic properties in the environment where eggs are developing such that they may be too cold or warm @geist2002.

@@

**BRYAN check the below**

The models presented in **hatchR** can be further customized beyond the use-cases provided above. For instance, while our models as presented are fit to predict hatch or emergence timing, they could be used to predict other developmental stages prior to the initation of external feeding like eye-up or embryological developmental stages @velsen1980 or **Some example from a non-salmonid, maybe initiation of pelagic-larval or riverine current dispersal**. Additionally, while `fit_model()` relies on non-linear modeling platform, `predict_phenology()` only requires users to pass a model expression, so if users preferred a different model fit than the non-linear option provided, they could pass a different expression into `predict_phenology()` for additional customization. Finally, while **hatchR** was designed specifically for fishes, we expect other poikilothermic organisms such as reptiles, amphibians, or invertebrates would develop in an exponential decay relationship and these developmental models could be extended to other organisms beyond fishes.
